The forestecology R package for fitting and assessing neighborhood models of the effect of interspecific competition on the growth of trees

### 2021-03-10 paper/forestecology.Rmd
3139 -> 3060. Gah!

keywords:
- forest ecology, interspecific competition, neigbhorhood competition, tree growth, R, ForestGEO, spatial cross-validation

@@

Repeat-censused forest plots offer excellent opportunities to test neighborhood models of the effect of competition on the growth of trees (@canham_neighborhood_2004). Neighborhood models of competition have been used to: test whether the species identity of a competitor matters (@uriarte_spatially_2004); measure species-specific competition coefficients (@das_effect_2012 @tatsumi_estimating_2016); test competing models to see what structures competitive interactions, e.g. traits or phylogeny (@allen_permutation_2020; @uriarte_trait_2010); and inform selective logging practices (@canham_neighborhood_2006). Although these are well-described methods, few methods are currently available for easy application. Here we address this in an R package. We largely follow the methods presented in @allen_permutation_2020. The package is written to model stem radial growth between two censuses based on neighborhood competition.

@allen_permutation_2020 considers the following model: Let $i = 1, \ldots, n_j$ index all $n_j$ trees of "focal" species group $j$; let $j = 1, \ldots, J$ index all $J$ focal species groups; and let $k = 1, \ldots, K$ index all $K$ "competitor" species groups. They use the following linear model the average annual growth in diameter at breast height (DBH) $y_{ij}$ (in centimeters/year) of the $i^{th}$ tree of focal species group $j$:

\begin{equation}
\label{eq:model}
y_{ij} = f(\vec{x}_{ij}) + \epsilon_{ij} = \beta_{0,j} + \beta_{\text{dbh},j} \cdot \text{dbh}_{ij} + \sum_{k=1}^{K} \lambda_{jk} \cdot \text{BA}_{ijk} + \epsilon_{ij}
\end{equation}

where $\beta_{0,j}$ is the diameter-independent growth rate for group $j$; $\text{dbh}_{ij}$ is the DBH the focal tree at the earlier census; $\beta_{\text{dbh},j}$ is the amount of the growth rate changed depending on diameter for group $j$;  $\text{BA}_{ijk}$ is the sum of the basal area of all trees of competitor species group $k$; $\lambda_{jk}$ is the change in growth for individuals of group $j$ from nearby competitors of group $k$; and $\epsilon_{ij}$ is a random error term distributed $\text{Normal}(0, \sigma^2)$. They estimate all parameters via Bayesian linear regression while exploiting Normal/Inverse Gamma conjugacy to derive closed-form solutions to all posterior distributions^[See S1 Appendix of @allen_permutation_2020, available at https://doi.org/10.1371/journal.pone.0229930.s004]. These closed-form solutions for the posterior distributions are in contrast to approximations of all posteriors via computationally expensive Markov Chain Monte Carlo algorithms.

In order to evaluate whether competitor species identity matters, @allen_permutation_2020 run a permutation test where a null hypothesis of no species grouping-specific effects of competition is assumed, and thus the species identity of all competitors can be permuted/shuffled:

\begin{eqnarray}
\label{eq:permutation-hypothesis-test}
&&H_0: \lambda_{jk} = \lambda_{j} \mbox{ for all } k = 1, \ldots, K\\
\text{vs.}&&H_A: \text{at least one } \lambda_{jk} \mbox{ is different}
\end{eqnarray}

Furthermore, in order to account for the spatial autocorrelation inherent to forest data in their estimates of out-of-sample model error, @allen_permutation_2020 use spatial cross-validation. Estimates of model error that do not account for this spatial dependency tend to underestimate the true model error [@roberts_cross-validation_2017].

We introduce the `forestecology` R package providing methods and data for forest ecology model fitting and assessment, available on CRAN (https://cran.r-project.org/web/packages/forestecology/index.html) and on GitHub (https://github.com/rudeboybert/forestecology). The package implements all aspects of the model in Equation \ref{eq:model}: model fitting and generating predicted values, evaluating the effect of competitor species identity using permutation tests, and evaluating model performance using spatial cross-validation.

The package designed with "tidy" design principles in mind [@wickham_welcome_2019]. Much like all `tidyverse` packages, `forestecology` is has verb-named functions that can be modularly composed using the pipe `%>%` operator to complete all the necessary steps in the analysis sequence [@bache_pipe_2020]. Furthermore, the inputs and outputs of most of our functions use the same "simple features for R" data structures from the `sf` package for standardized and `tidyverse`-friendly wrangling and visualizing of spatial data [@pebesma_simple_2018]

Currently the package only implements the Bayesian linear regression model detailed in Equation \ref{eq:model}. As we demonstrate in Section \ref{model-fit-predict} however, the fitting of this model is self-contained in a single function `comp_bayes_lm()` which returns an object of S3 class type `comp_bayes_lm`. This class has generic methods implemented to print, make predictions using, and plot all results. Therefore the package can be modularly extended to fit other models as long as they are coded similarly as `comp_bayes_lm()` and have equivalent generic methods implemented.

@@

We present a case-study of the `forestecology` package's functionality on data from the Smithsonian Conservation Biology Institute (SCBI) large forest dynamics plot in Front Royal, VA, USA, part of the ForestGEO global network of research sites [@bourg_initial_2013, @andersonteixeira_ctfs-forestgeo_2015] [@bourg_initial_2013]. The 25.6 ha (640 x 400 m) plot is located at the intersection of three of the major physiographic provinces of the eastern US---the Blue Ridge, Ridge and Valley, and Piedmont provinces---and is adjacent to the northern end of Shenandoah National Park.

The package has the following ecological goals: to evaluate i) the effect of competitor species identity using permutation tests and ii) model performance using spatial cross-validation. We outline a basic analysis sequence comprising of the following four main steps:

@@

The first step is to compute the growth of trees using data from two censuses. `compute_growth()` computes the average annual growth based on census data that roughly follows ForestGEO standards. Despite such standards, minor variations will still exist between sites, thereby necessitating some data wrangling. For example, the SCBI site records all DBH values in millimeters [@bourg_initial_2013], whereas the Michigan Big Woods site records them in centimeters [@allen_michigan_2020].

We first load both 2008 and 2014 SCBI census `.csv` files as they existed on GitHub on 2020/11/20 and perform some data wrangling to both data sets [@gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020]. We then only consider a 9 ha subsection of the 25.6 ha of the SCBI site in order to speed up computation for this example: `gx` from 0--300 instead of 0--400 and `gy` from 300--600 instead of 0--640.

@@

These two data frames are then used as inputs to `compute_growth()`, along with the `id` argument specifying the variable that uniquely identifies each tree-stem. Note that we also discard all resprouts in the later census with `code == R`, since we are only interested in the growth of surviving, and not resprouted, stems.

```{r scbi-compute-growth}
growth_scbi <-
  compute_growth(
    census_1 = census_2013_scbi,
    census_2 = census_2018_scbi %>% filter(!str_detect(codes, "R")),
    id = "stemID"
  )
growth_scbi
```

The output `growth_scbi` is a data frame of class `sf` that includes variables the average annual `growth` in DBH (cm $\cdot$ y^-1^) for all stems that were alive at both time points, the `sf` package's encoding of geolocations of `geometry` type `<POINT>`, and the species variable `sp` converted to a factor. Given that `growth_scbi` is of class `sf`, it can be easily plotted in `ggplot2` using `geom_sf()` as seen in Figure \ref{fig:scbi-trees}.

```{r scbi-trees, out.width="66%", fig.cap="Compute growth of trees based on census data: Map with growth of a random sample of 500 trees from a 9 ha subsection of the Smithsonian Conservation Biology Institute (SCBI) forest plot."}
ggplot() +
  geom_sf(data = growth_scbi %>% sample_n(500), aes(size = growth)) + 
  scale_size_binned(limits = c(0.1, 1))
```

@@

The next step is to add spatial information to `growth_scbi`. The first element we add is a "buffer region" to the periphery of the study region. Since some of our model's explanatory variables are cumulative, we must ensure that all trees being modeled are not biased to have different neighbor structures. This is of concern for trees at the boundary of the study region whose neighbors will not all be included in the censused stems. To account for such edge effects only trees that are not part of this buffer region, i.e. are part of the interior of the study region, will have their growth modeled [@waller_applied_2004].

Our model of interspecific competition relies on a spatial definition of who competitor trees are: all trees within a distance `comp_dist` of a focal tree. In our case we set `comp_dist` to 7.5m, a value informed by other studies [@canham_neighborhood_2004, @uriarte_spatially_2004, @canham_neighborhood_2006]. We use `comp_dist` and a manually constructed `sf` representation of the study region's boundary as inputs to `add_buffer_variable()` to add a `buffer` boolean variable to `growth_scbi`. All trees with `buffer` equal to `FALSE` will be our focal trees whose growth will be modeled, whereas those with `TRUE` will only act as competitor trees.

@@

The second element of spatial information we add are blocks corresponding to folds of a spatial cross-validation algorithm. Conventional cross-validation algorithms assign observations to folds by randomly resampling individual observations that are assumed independent. In the case of forest census data however, observations exhibit spatial autocorrelation. We therefore incorporate this dependence into the cross-validation algorithm by resampling spatial blocks of trees [@roberts_cross-validation_2017 @pohjankukka_estimating_2017].

We first manually an `sf` object defining four folds that partition the study region. We then use the output of the `spatialBlock()` function from the `blockCV` package to associate each tree in `growth_scbi` to the correct `foldID` [@valavi_blockcv_2019].^[In the Supporting Information we present an example where the folds themselves are created automatically, as opposed to manually as in the example.] This `foldID` variable will be used in Section \ref{spatial-cross-validation}.

@@

# Associate each observation to a fold
spatial_block_scbi <- spatialBlock(
  speciesData = growth_scbi, k = n_fold, selection = "systematic", 
  blocks = blocks_scbi, showBlocks = FALSE, verbose = FALSE
)

@@

```{r scbi-spatial-information, out.width="66%", fig.cap="Add spatial information: Buffer region and spatial cross-validation blocks (1 through 4). The location of each tree is marked with an integer indicating its fold, with folds delineated with solid lines. The color of each digit indicates whether the tree is part of the buffer region (and thus will only be considered as a competitor tree in our model) or is part of the interior of the study region (and thus is a focal tree whose growth is of modeled interest)."}
ggplot() +
  geom_sf(data = blocks_scbi, fill = "transparent", linetype = "dashed") +
  geom_sf_text(data = growth_scbi %>% sample_n(1000), 
               aes(label = foldID, col = buffer))
```

@@

The next step is to identify all focal trees and their corresponding competitor trees. More specifically, identify all trees that are not part of the buffer region, have a valid `growth` measurement, and have at least one neighbor within `r comp_dist`m. We do this using `create_focal_vs_comp()`, which takes the previously detailed arguments `comp_dist` and `id` as well as the `sf` representation of the spatial cross-validation blocks and returns a new data frame `focal_vs_comp_scbi`.

@@

We observe 4 variables describing 20 competitor trees: their unique tree-stem ID, their distance to the focal tree (all $\leq$ `r comp_dist`), their species, and their basal area (in m$^2$) calculated as $\frac{\pi \times (\text{DBH/2})^2}{10000}$ for $DBH$ in cm from the earlier census. Saving competitor information in list-column format minimizes redundancy since we do not need to repeat information on the focal tree 20 times. We visualize the spatial distribution of these trees in Figure \ref{fig:scbi-focal-vs-comp-map}.

```{r scbi-focal-vs-comp-map, out.width="66%", echo = FALSE, fig.cap="Identify all focal and corresponding competitor trees: The dashed circle extends 7.5m away from the focal tree 4 while all 20 competitor trees are within this circle."}
comp_radius <- growth_scbi %>% 
  filter(stemID == 4) %>% 
  st_buffer(dist = comp_dist)

@@

The final step is to fit the competition Bayesian linear regression model for tree growth outlined in Equation \ref{eq:model} using `comp_bayes_lm()`. This function has an option to specify prior distributions on all parameters of interest, chosen here to be the defaults detailed in `?comp_bayes_lm`.

@@

The resutling `comp_bayes_lm_scbi` is an object of S3 class type `comp_bayes_lm` containing the posterior values of all parameters of our model. Furthermore, this class of object includes generics for three methods. First, the generic for `print()` displays the names of all prior and posterior parameters along with the model formula:

@@

Next, the generic for `predict()` takes the posterior parameter values in `comp_bayes_lm_scbi` and a `newdata` data frame and outputs a vector `growth_hat` of predicted values $\widehat{y}$ of the DBH for each focal tree computed from the posterior predictive distribution.

@@

We then compare the observed and predicted growths to compute the root mean squared error (RMSE) of our model fit.

@@

```{r scbi-posterior-viz, out.width="100%", fig.cap= "Fit model: Posterior distributions of all parameters. For compactness we include only three species.", fig.width = 8, fig.height = 6}
# Plot posteriors for only a subset of species
sp_to_plot <- c("litu", "quru", "cagl")

@@

For many users the visualizations relating to $\lambda_{j,k}$ will be of particular interest as they provide insight into species competitive interactions where negative values indicate a competitor species which slows the growth of a focal species. Here, for example, we see that tulip poplars (`litu`) have a strong negative effect on the growth of conspecifics but relatively lesser effect on neighbors of the other two species.

@@

To evaluate the effect of competitor species identity, we use the above four steps along with a permutation test. Under a null hypothesis where competitor species identity does not matter, we permute the competitor species identity within each focal tree, compute the RMSE test statistic, repeat this process several times to construct a null distribution, and finally compare it to the observed RMSE to assess significance. Going back to our example in Section \ref{focal-vs-comp} of focal tree with `focal_ID` 4 and its 20 competitors, the permutation test randomly resamples only the `comp_sp` variable without replacement, leaving all other variables intact. The resampling without replacement is nested within each focal tree in order to preserve neighborhood structure. We once again use `comp_bayes_lm()`, but this time setting `run_shuffle = TRUE`.

@@

focal_vs_comp_scbi <- focal_vs_comp_scbi %>%
  mutate(
    growth_hat_shuffle = predict(comp_bayes_lm_scbi_shuffle, 
                                 newdata = focal_vs_comp_scbi)
  )
```
```{r}
model_rmse_shuffle <- focal_vs_comp_scbi %>%
  rmse(truth = growth, estimate = growth_hat_shuffle) %>%
  pull(.estimate)
model_rmse_shuffle
```

@@

To evaluate model performance, we use spatial cross-validation. The model fits and predictions in Section \ref{model-fit-predict} use the same data to both fit the model and to assess the model's performance. Given the inherent spatial-autocorrelation of our data, this can potentially lead to potentially overfit models [@roberts_cross-validation_2017]. To mitigate this potential overfitting, we use the spatial cross-validation blocking scheme encoded in the `foldID` variable from Section \ref{spatial-information} and visualized in Figure \ref{fig:scbi-spatial-information}.

At each iteration of our cross-validation, one fold will act as the test set with the remaining three acting as the training set. We fit the model to all focal trees in the training set, apply the model to all focal trees in the test set, compute predicted values, and compute the RMSE. Furthermore, to maintain spatial independence between the test and training set, a "fold buffer" that extend outwards from the boundary of the test set is computed; all trees falling within this fold buffer are excluded from the training set (see Figure \ref{fig:scbi-spatial-cross-validation-schematic}).

```{r scbi-spatial-cross-validation-schematic, out.width="66%", fig.cap= "Schematic of spatial cross-validation: Using the k = 1 fold (bottom-left) as the test set, k = 2 through 4 as the training set, along with a fold buffer.", echo = FALSE}
fold_number <- 1

@@

This is repeated for each of the four folds acting as the test set and then the four resulting RMSE's are averaged to provide a single estimate of model error. This algorithm is implemented in `run_cv()`, which acts as a wrapper function to both `comp_bayes_lm()` that fits the model and `predict()` that returns predicted values.

@@

The resulting RMSE of `r model_rmse_cv` computed using cross-validation is `r ifelse(model_rmse_cv > model_rmse, "larger", "smaller")` than the earlier RMSE of `r model_rmse`, suggesting that models that do not take spatial autocorrelation account generate moel error estimates that are overly optimistic. In our case, RMSE values that are too low.

@@

`run_cv()` also accepts the `run_shuffle` argument in order to permute competitor species identity as described in Section \ref{permutation-test}. Figure \ref{fig:scbi-simulation} compares model performance for 49 permutations of competitor species and calculating RMSE, both with and without cross-validation. Without cross-validation, competitor species identity did matter as the observed RMSE was lower than the permutation null distribution of RMSE. However, once we incorporate spatial cross-validation, this improvement disappears. These results suggest that in this 9 ha subplot of the SCBI plot competitive interactions do not depend on the identity of the competitor, which is the opposite of what has been observed in other locations [@allen_permutation_2020 @uriarte_spatially_2004]. Furthermore, the larger cross-validated RMSE's are indicative of the importance of being vigilant against model overfitting.

@@

The `forestecology` package provides an accessible way to fit and test models of neighborhood competition. While the package is designed with ForestGEO plot data in mind, we envision that it can be modified to work on i) any single large, mapped forest plot in which at least two measurements of each individual have been taken, e.g. the US Forest Service Forest Inventory and Analysis plots, or ii) more generally to model interactions of any community of mapped sessile organisms [@smith_forest_2002]. In future versions of `forestecology` we also hope to include models that account for tree mortality in addition to tree growth. The package follows the guidelines for `tidy` data, leverages the `sf` package for spatial data, and S3 open-oriented model structure. We hope that the package will increase the use of neighborhood competition models to better understand what structures plant competition.

@@

We intend to archive all data and source code for this manuscript on GitHub at https://github.com/rudeboybert/forestecology. This repository will be archived on Zenodo upon acceptance. The example Smithsonian Conservation Biology Institute census data used are available on GitHub at https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_main_census/data/census-csv-files and are archived on Zenodo at https://doi.org/10.5281/zenodo.2649301 [@gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020].

>>>>

keywords:
- forest ecology, interspecific competition, neighborhood competition, tree growth, R, ForestGEO, spatial cross-validation

@@

Repeat-censused forest plots offer excellent opportunities to test neighborhood models of the effect of competition on the growth of trees (@canham_neighborhood_2004). Neighborhood models of competition have been used to: test whether the species identity of a competitor matters (@uriarte_spatially_2004); measure species-specific competition coefficients (@das_effect_2012 @tatsumi_estimating_2016); test competing models to see what structures competitive interactions, e.g. traits or phylogeny (@allen_permutation_2020; @uriarte_trait_2010); and inform selective logging practices (@canham_neighborhood_2006). Although these are well-described methods, few methods are currently available for easy application.

We address this shortcoming with the `forestecology` R package providing methods and data for forest ecology model fitting and assessment available on CRAN (https://cran.r-project.org/web/packages/forestecology/index.html) and on GitHub (https://github.com/rudeboybert/forestecology). The package is written to model stem radial growth between two censuses based on neighborhood competition largely following the methods in @allen_permutation_2020.

Let $i = 1, \ldots, n_j$ index all $n_j$ trees of "focal" species group $j$; let $j = 1, \ldots, J$ index all $J$ focal species groups; and let $k = 1, \ldots, K$ index all $K$ "competitor" species groups. The average annual growth in diameter at breast height (DBH) $y_{ij}$ (in centimeters/year) of the $i^{th}$ tree of focal species group $j$ is modeled as

\begin{equation}
\label{eq:model}
y_{ij} = \beta_{0,j} + \beta_{\text{dbh},j} \cdot \text{dbh}_{ij} + \sum_{k=1}^{K} \lambda_{jk} \cdot \text{BA}_{ijk} + \epsilon_{ij}
\end{equation}

where $\beta_{0,j}$ is the diameter-independent growth rate for group $j$; $\text{dbh}_{ij}$ is the DBH of the focal tree at the earlier census for group $j$ and $\beta_{\text{dbh},j}$ quantifies its corresponding growth rate change;  $\text{BA}_{ijk}$ is the sum of the basal area of all trees of competitor species group $k$ and $\lambda_{jk}$ quantifies the corresponding change in growth for individuals of group $j$ from these competitors; and $\epsilon_{ij}$ is a random error term distributed $\text{Normal}(0, \sigma^2)$. They estimate all parameters via Bayesian linear regression while exploiting Normal/Inverse Gamma conjugacy to derive closed-form solutions to all posterior distributions^[See S1 Appendix of @allen_permutation_2020, available at https://doi.org/10.1371/journal.pone.0229930.s004]. These closed-form solutions are in contrast to approximations of all posteriors via computationally expensive Markov Chain Monte Carlo algorithms.

To evaluate whether competitor species identity matters, @allen_permutation_2020 run a permutation test where a null hypothesis of no species grouping-specific effects of competition is assumed, and thus the species identity of all competitors can be permuted:

\begin{eqnarray}
\label{eq:permutation-hypothesis-test}
&&H_0: \lambda_{jk} = \lambda_{j} \mbox{ for all } k = 1, \ldots, K\\
\text{vs.}&&H_A: \text{at least one } \lambda_{jk} \mbox{ is different} \nonumber
\end{eqnarray}

Furthermore, to account for the spatial autocorrelation of forestry data in their estimates of out-of-sample model error, @allen_permutation_2020 use spatial cross-validation. Estimates of model error that do not account for this spatial dependency tend to underestimate the true model error [@roberts_cross-validation_2017].

The package is designed with "tidy" design principles in mind [@wickham_welcome_2019]. Much like all `tidyverse` packages, `forestecology` has verb-named functions that can be modularly composed using the pipe `%>%` operator to sequentially complete all the necessary analysis steps [@bache_pipe_2020]. Furthermore, the inputs and outputs of most of our functions use the same "simple features for R" data structures from the `sf` package, a package for standardized and `tidyverse`-friendly wrangling and visualizing of spatial data [@pebesma_simple_2018]

Currently the package only implements the Bayesian linear regression model detailed in Equation \ref{eq:model}. As we demonstrate in Section \ref{model-fit-predict} however, the fitting of this model is self-contained in a single function `comp_bayes_lm()` which returns an object of S3 class type `comp_bayes_lm`. This class has generic methods implemented to print, make predictions, and plot all results. Therefore the package can be modularly extended to fit other models as long as they are coded similarly to `comp_bayes_lm()` and have equivalent generic methods implemented.

@@

We present a case-study of `forestecology`'s functionality on data from the Smithsonian Conservation Biology Institute (SCBI) large forest dynamics plot in Front Royal, VA, USA, part of the ForestGEO global network of research sites [@bourg_initial_2013, @andersonteixeira_ctfs-forestgeo_2015] [@bourg_initial_2013]. The 25.6 ha (640 x 400 m) plot is located at the intersection of three of the major physiographic provinces of the eastern US---the Blue Ridge, Ridge and Valley, and Piedmont provinces---and is adjacent to the northern end of Shenandoah National Park.

The package has the following ecological goals: to evaluate i) the effect of competitor species identity using permutation tests and ii) model performance using spatial cross-validation. We outline the four step basic analysis sequence:

@@

The first step is to compute the growth of trees using data from two censuses. `compute_growth()` computes the average annual growth based on census data that roughly follows ForestGEO standards. Despite such standards, minor variations will still exist between sites, thereby necessitating some data wrangling. For example, the SCBI site records all DBH values in millimeters [@bourg_initial_2013], whereas the Michigan Big Woods site used in @allen_permutation_2020 records them in centimeters [@allen_michigan_2020].

We first load both 2008 and 2014 SCBI census `.csv` files as they existed on GitHub on 2020/11/20 and perform minor data wrangling to both data sets [@gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020]. We then only consider a 9 ha subsection of the 25.6 ha of the SCBI site in order to speed up computation for this example: `gx` from 0--300 instead of 0--400 and `gy` from 300--600 instead of 0--640.

@@

These two data frames are then used as inputs to `compute_growth()`, along with the `id` argument specifying the variable that uniquely identifies each tree-stem. Note that we also discard all resprouts with `code == R` in the later census, since we are only interested in the growth of surviving, and not resprouted, stems.

```{r scbi-compute-growth}
growth_scbi <-
  compute_growth(
    census_1 = census_2013_scbi,
    census_2 = census_2018_scbi %>% filter(!str_detect(codes, "R")),
    id = "stemID"
  )
growth_scbi %>% 
  select(stemID, sp, dbh1, dbh2, growth, geometry)
```

The output `growth_scbi` is a data frame of class `sf` that includes among other variables the species variable `sp` converted to a factor, the average annual `growth` in DBH (cm $\cdot$ y^-1^) for all stems that were alive at both time points, and the `sf` package's encoding of geolocations of `geometry` type `<POINT>`. Given that `growth_scbi` is of class `sf`, it can be easily plotted in `ggplot2` using `geom_sf()` as seen in Figure \ref{fig:scbi-trees}.

```{r scbi-trees, out.width="66%", fig.cap="Compute growth of trees based on census data. A map of the growth of a random sample of 500 trees from a 9 ha subsection of the Smithsonian Conservation Biology Institute (SCBI) forest plot."}
ggplot() +
  geom_sf(data = growth_scbi %>% sample_n(500), aes(size = growth)) + 
  scale_size_binned(limits = c(0.1, 1))
```

@@

We then add spatial information to `growth_scbi`. We first add a "buffer region" to the periphery of the study region. Since some of our model's explanatory variables are cumulative, we must ensure that all trees being modeled are not biased to have different neighbor structures. This is of concern for trees at the boundary of the study region who will not have all their neighbors included in the census stems. To account for such edge effects, only trees that are not part of this buffer region, i.e. are part of the interior of the study region, will have their growth modeled [@waller_applied_2004].

Our model of interspecific competition relies on a spatial definition of who competitor trees are: all trees within a distance `comp_dist` of a focal tree. We set `comp_dist` to 7.5m, a value informed by other studies [@canham_neighborhood_2004, @uriarte_spatially_2004, @canham_neighborhood_2006]. We use `comp_dist` and a manually constructed `sf` representation of the study region's boundary as inputs to `add_buffer_variable()` to add a `buffer` boolean variable to `growth_scbi`. All trees with `buffer` equal to `FALSE` will be our focal trees whose growth will be modeled, whereas those with `TRUE` will only act as competitor trees.

@@

The second element of spatial information we add are blocks corresponding to folds of a spatial cross-validation algorithm. Conventional cross-validation algorithms assign individual observations to folds by randomly resampling them all while assuming they are statistically independent. In the case of forest census data however, observations exhibit spatial autocorrelation. We therefore incorporate this dependence into the cross-validation algorithm by resampling spatial blocks of trees [@roberts_cross-validation_2017 @pohjankukka_estimating_2017].

We first manually define an `sf` object defining four folds that partition the study region. We then use the output of the `spatialBlock()` function from the `blockCV` package to associate each tree in `growth_scbi` to the correct `foldID` [@valavi_blockcv_2019].^[In the Supporting Information we present an example where the folds themselves are created automatically, as opposed to manually as in the example.] This `foldID` variable will be used in Section \ref{spatial-cross-validation}.

@@

# Associate each observation to a fold
spatial_block_scbi <- 
  spatialBlock(speciesData = growth_scbi, k = n_fold, 
               selection = "systematic", blocks = blocks_scbi, 
               showBlocks = FALSE, verbose = FALSE)

@@

```{r scbi-spatial-information, out.width="66%", fig.cap="Add spatial information. A buffer region and spatial cross-validation blocks 1 througth 4. The location of each tree is marked with its fold number where the folds are delineated with solid lines. The color of each digit indicates whether the tree is part of the buffer region (thus will only be considered as a competitor tree) or is part of the interior of the study region (thus is a focal tree whose growth is of modeled interest)."}
ggplot() +
  geom_sf(data = blocks_scbi, fill = "transparent", linetype = "dashed") +
  geom_sf_text(data = growth_scbi %>% sample_n(1000), 
               aes(label = foldID, col = buffer))
```

@@

We then identify all focal trees and their corresponding competitor trees. More specifically, identify all trees that are not part of the buffer region, have a valid `growth` measurement, and have at least one neighbor within `r comp_dist`m. We do this using `create_focal_vs_comp()`, which takes the previously detailed `comp_dist` and `id` arguments as well as the `sf` representation of the spatial cross-validation blocks and returns a new data frame `focal_vs_comp_scbi`.

@@

We observe 4 variables describing 20 competitor trees: the unique tree-stem ID, the distance to the focal tree (all $\leq$ `r comp_dist`), the species, and the basal area (in m$^2$) calculated as $\frac{\pi \times (\text{DBH/2})^2}{10000}$ for the DBH in cm from the earlier census. Saving competitor information in list-column format minimizes redundancy since we do not need to repeat information on the focal tree 20 times. We visualize the spatial distribution of these trees in Figure \ref{fig:scbi-focal-vs-comp-map}.

```{r scbi-focal-vs-comp-map, out.width="66%", echo = FALSE, fig.cap="Identify all focal and corresponding competitor trees. The dashed circle extends 7.5m away from the focal tree 4 while all 20 competitor trees are within this circle."}
comp_radius <- growth_scbi %>% 
  filter(stemID == 4) %>% 
  st_buffer(dist = comp_dist)

@@

Lastly, we then fit the competition Bayesian linear regression model for tree growth outlined in Equation \ref{eq:model} using `comp_bayes_lm()`. This function has an option to specify prior distributions of all parameters of interest, chosen here to be the defaults detailed in `?comp_bayes_lm`.

@@

The resulting`comp_bayes_lm_scbi` is an object of S3 class type `comp_bayes_lm` containing the posterior values of all parameters. Furthermore, this class of object includes generics for three methods. First, the generic for `print()` displays the names of all prior and posterior parameters and the model formula:

@@

Next, the generic for `predict()` takes the posterior parameter values in `comp_bayes_lm_scbi` and a `newdata` data frame, and outputs a vector `growth_hat` of predicted values $\widehat{y}$ of the DBH computed from the posterior predictive distribution.

@@

We can now compare the observed and predicted growths to compute the root mean squared error (RMSE) of our model fit.

@@

```{r scbi-posterior-viz, out.width="100%", fig.cap= "Fit model. Posterior distributions of all parameters. For compactness we include only three species.", fig.width = 8, fig.height = 6}
# Plot posteriors for only a subset of species
sp_to_plot <- c("litu", "quru", "cagl")

@@

For many users the visualizations relating to $\lambda_{j,k}$ will be of particular interest as they provide insight into species competitive interactions, where negative values indicate a competitor species which slows the growth of a focal species. Here, for example, we see that tulip poplars (litu) have a strong negative effect on the growth of conspecifics but relatively lesser effect on pignut hickory (cagl) and red oak (quru) neighbors.

@@

To evaluate the effect of competitor species identity, we use the above four steps along with the permutation test in Equation \ref{eq:permutation-hypothesis-test}. Under a null hypothesis where competitor species identity does not matter, we can permute the competitor species identities within each focal tree, compute the RMSE test statistic, repeat this process several times to construct a null distribution, and finally compare it to the observed RMSE to assess significance. Going back to our example in Section \ref{focal-vs-comp} of focal tree with `focal_ID` 4 and its 20 competitors, the permutation test only randomly resamples the `comp_sp` variable without replacement, leaving all other variables intact. This resampling is nested within each focal tree in order to preserve neighborhood structure. We perform this permutation test, we once again use `comp_bayes_lm()` but by setting `run_shuffle = TRUE`.

@@

focal_vs_comp_scbi <- focal_vs_comp_scbi %>%
  mutate(growth_hat_shuffle = predict(comp_bayes_lm_scbi_shuffle, 
                                 newdata = focal_vs_comp_scbi))
```
```{r}
model_rmse_shuffle <- focal_vs_comp_scbi %>%
  rmse(truth = growth, estimate = growth_hat_shuffle) %>%
  pull(.estimate)
model_rmse_shuffle
```

@@

To evaluate model performance, we use spatial cross-validation. The example model fit in Section \ref{model-fit-predict} uses the same data to both fit the model and assess its performance. Given the inherent spatial-autocorrelation of our data, this can potentially lead to overfit models [@roberts_cross-validation_2017]. To mitigate the risk of overfitting, we make use of the spatial cross-validation blocking scheme encoded in the `foldID` variable from Section \ref{spatial-information} and visualized in Figure \ref{fig:scbi-spatial-information}.

At each iteration of our cross-validation, one fold will act as the test set and the remaining three acting as the training set. We fit the model to all focal trees in the training set, apply the model to all focal trees in the test set, compute predicted values, and compute the RMSE. Furthermore, to maintain spatial independence between the test and training set, a "fold buffer" that extend `r comp_dist`m outwards from the boundary of the test set is computed; all trees falling within this fold buffer are excluded from the training set (see Figure \ref{fig:scbi-spatial-cross-validation-schematic}).

```{r scbi-spatial-cross-validation-schematic, out.width="66%", fig.cap= "Schematic of spatial cross-validation. Using the k = 1 fold (bottom-left) as the test set, k = 2 through 4 as the training set, along with a \"fold buffer\" extending outwards from the test set to maintain spatial independence between it and the training set.", echo = FALSE}
fold_number <- 1

@@

This is repeated for each of the four folds acting as the test set, then the four resulting RMSE's are averaged to provide a single estimate of model error. This algorithm is implemented in `run_cv()`, which acts as a wrapper function to both `comp_bayes_lm()` that fits the model and `predict()` that returns predicted values.

@@

The resulting RMSE of `r model_rmse_cv` computed using cross-validation is `r ifelse(model_rmse_cv > model_rmse, "larger", "smaller")` than the earlier RMSE of `r model_rmse`, suggesting that models that do not take spatial autocorrelation account generate model error estimates that are overly optimistic. In our case, RMSE values that are too low.

@@

`run_cv()` also accepts the `run_shuffle` argument in order to permute competitor species identity as described in Section \ref{permutation-test}. Figure \ref{fig:scbi-simulation} compares model performance for 49 permutations of competitor species and RMSE calculations, both with and without cross-validation. Without cross-validation, competitor species identity did matter as the observed RMSE was significantly lower than the permutation null distribution of RMSE. However, once we incorporate spatial cross-validation, this improvement disappears. These results suggest that in this 9 ha subplot of the SCBI plot competitive interactions do not depend on the identity of the competitor, which is the opposite of what has been observed in other locations [@allen_permutation_2020 @uriarte_spatially_2004]. Furthermore, the larger cross-validated RMSE's are indicative of the importance of being vigilant against potential model overfitting.

@@

The `forestecology` package provides an accessible way to fit and test models of neighborhood competition. While the package is designed with ForestGEO plot data in mind, we envision that it can be modified to work on any single large, mapped forest plot in which at least two measurements of each individual have been taken, e.g. the US Forest Service Forest Inventory and Analysis plots. Furthemore, we also envision the package being used more generally to model interactions of any community of mapped sessile organisms [@smith_forest_2002]. In future versions of `forestecology` we also hope to include models that account for tree mortality in addition to tree growth. The package follows the `tidy` data design principles, leverages the `sf` package for spatial data, and S3 open-oriented model implementation structure. We hope that the package will increase the use of neighborhood competition models to better understand what structures plant competition.

@@

We intend to archive all data and source code for this manuscript on GitHub at https://github.com/rudeboybert/forestecology. This repository will be archived on Zenodo upon acceptance. The example Smithsonian Conservation Biology Institute census data are available on GitHub at https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_main_census/data/census-csv-files and are archived on Zenodo at https://doi.org/10.5281/zenodo.2649301 [@gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020].
